The title

### 2016-10-04 analysis/paper/paper.Rmd
updated references

# Dennell is a major commentator recently
Dennell is one of the more recent critics of the concept of the Movius Line. In his analysis of Movius’ publications, he describes their views as “backwards”, “ancient” and “eurocentric” [@dennell2016life]. The drawing of a line to mark difference in stone tool production between SE and SW/W Asia neglects the variety and complexity of lithic assemblages on either side of the line, leaving East Asia in a minor position in human evolution. According to Dennell, none of the material found by Movius and his colleagues has a stratigraphic context and they failed to identify a sequence of four terraces along the Irrawaddy River from the Middle to Upper Pleistocene. Furthermore, because of the finding context, his finds cannot be connected to other Middle Pleistocene Acheulean assemblages in Southwest Asia [@dennell2016life].
In contrast to Dennell’s views, Norton et al. (2006) reworked the concept of the Movius Line by incorporating existing issues with the “traditional” Movius Line. The three main characteristics of the “Movius Line sensu lato” are the lower frequency of handaxe bearing sites in East Asia compared to Africa and India; the much lower percentage of bifacially made tools in East Asian assemblages and the morphological similarities to Acheulean artefacts [@norton2006middle; @norton2008movius].
Instead of assigning stone tools to certain technologies made by different groups of early humans, and making implications about their abilities to produce stone tools, acknowledging the spatial and temporal diversity of lithic records in Eurasia is necessary [@dennell2016life]. Reasons for the lack of Acheulean technology are investigated in modern quaternary science [@brumm2012biface; @dennell2016life; @norton2008movius; @norton2006middle; @petraglia2008large; @schick1994movius], including constraints on raw material, demographic and social transmission, environmental changes and dispersal routes.
The use of stone tools may have been influenced by certain situations, tasks or individuals. Consequently, different members of a group or different groups could have used bifaces rarely, never or only at specific events or times. Thus, the absence of bifaces or an item in general does not mean that the hominin living in this area at this time was not able to produce such a tool [@dennell2016life].

@@

Movius (1944) stated that the raw material used east of the Movius Line was often of low quality quartz and quartzite that would prevent the production of the same kind of bifaces [@dennell2016life; @lycett2010movius]. However, handaxes from Zhoukoudian Locality 1, Chongokni and Kumpari (China) are bifacially worked and are made of quartz and quartzite river cobbles from this area [@lycett2010movius].

@@

The dispersal route of early modern humans from Africa to Asia and Australia is still debated among quaternary scientists. Based on GIS analyses, Field & Lahr (2006) and Field et al. (2007) identified possible routes from Africa leading eastwards along the coasts to Asia and eventually via the Sunda Shelf to Australia. This “Southern Dispersal Route” is dated to around 75-60ka [@field2005assessment; @field2007southern; @clarkson2012continuity; @macaulay2005single]. This route matches with some Middle Pleistocene archaeological sites in Africa, India (e.g. Borra, the Kokan Complex and the Hiran Valley) and SE Asia adjacent to coastal regions. Of particular importance are the deposits at the Hiran Valley since they have been dated to 56-69ka, matching the proposed time for the Southern Dispersal. A dispersal along the coasts and rivers seems reasonable because they provide sufficient resources. However, these first temporary settlements would not necessarily leave behind an archaeological record due to the rising sea level and the subsequent flooding of the Sunda shelf because of environmental changes QUELLE.

>>>>

# Dennell is a major commentator recently
Dennell is one of the more recent critics of the concept of the Movius Line. In his analysis of Movius’ publications, he describes their views as “backwards”, “ancient” and “eurocentric” [@dennell2016life]. The drawing of a line to mark difference in stone tool production between SE and SW/W Asia neglects the variety and complexity of lithic assemblages on either side of the line, leaving East Asia in a minor position in human evolution. According to Dennell, none of the material found by Movius and his colleagues has a stratigraphic context and they failed to identify a sequence of four terraces along the Irrawaddy River from the Middle to Upper Pleistocene. Furthermore, because of the finding context, his finds cannot be connected to other Middle Pleistocene Acheulean assemblages in Southwest Asia [@dennell2016life].
In contrast to Dennell’s views, Norton et al. [-@norton2006middle] reworked the concept of the Movius Line by incorporating existing issues with the “traditional” Movius Line. The three main characteristics of the “Movius Line sensu lato” are the lower frequency of handaxe bearing sites in East Asia compared to Africa and India; the much lower percentage of bifacially made tools in East Asian assemblages and the morphological similarities to Acheulean artefacts [@norton2006middle; @norton2008movius].
Instead of assigning stone tools to certain technologies made by different groups of early humans, and making implications about their abilities to produce stone tools, acknowledging the spatial and temporal diversity of lithic records in Eurasia is necessary [@dennell2016life]. Reasons for the lack of Acheulean technology are investigated in modern quaternary science [@brumm2012biface; @dennell2016life; @norton2008movius; @norton2006middle; @petraglia2008large; @schick1994movius], including constraints on raw material, demographic and social transmission, environmental changes and dispersal routes.
The use of stone tools may have been influenced by certain situations, tasks or individuals. Consequently, different members of a group or different groups could have used bifaces rarely, never or only at specific events or times. Thus, the absence of bifaces or an item in general does not mean that the hominin living in this area at this time was not able to produce such a tool [@dennell2016life].

@@

Movius [-@movius1944early] stated that the raw material used east of the Movius Line was often of low quality quartz and quartzite that would prevent the production of the same kind of bifaces [@dennell2016life; @lycett2010movius]. However, handaxes from Zhoukoudian Locality 1, Chongokni and Kumpari (China) are bifacially worked and are made of quartz and quartzite river cobbles from this area [@lycett2010movius].

@@

The dispersal route of early modern humans from Africa to Asia and Australia is still debated among quaternary scientists. Based on GIS analyses, Field & Lahr [-@field2005assessment] and Field et al. [-@field2007southern] identified possible routes from Africa leading eastwards along the coasts to Asia and eventually via the Sunda Shelf to Australia. This “Southern Dispersal Route” is dated to around 75-60ka [@field2005assessment; @field2007southern; @clarkson2012continuity; @macaulay2005single]. This route matches with some Middle Pleistocene archaeological sites in Africa, India (e.g. Borra, the Kokan Complex and the Hiran Valley) and SE Asia adjacent to coastal regions. Of particular importance are the deposits at the Hiran Valley since they have been dated to 56-69ka, matching the proposed time for the Southern Dispersal. A dispersal along the coasts and rivers seems reasonable because they provide sufficient resources. However, these first temporary settlements would not necessarily leave behind an archaeological record due to the rising sea level and the subsequent flooding of the Sunda shelf because of environmental changes QUELLE.
